Eco-evolutionary dynamics of finite populations from first principles

### 2023-05-01 ind_files/03_BD_processes.tex
Correct minor typos

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' (Also see equation \eqref{nD_stoch_RM_noise_term_alt_representation} in Appendix \ref{App_alternative_representation} for a more elegant representation of these noise terms as an integral against a single Wiener process). For the case $m=2, \mu = 0$ (two interacting types of individuals with no mutations between types), after using the representation of noise terms presented in Appendix \ref{App_alternative_representation} and letting $p = p_1$, equation \eqref{nD_eqn_for_frequencies} becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
In this equation, if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. To see this, one can observe that the RHS of equation \eqref{two_species_stoch_RM} identically vanishes at both $p=0$ and $p=1$, since every term on the RHS contains the product $p(1-p)$. Further, since there is no mutation between types, $p=0$ (extinction of type 1 organisms) and $p=1$ (extinction of type 2 organisms) are both `absorbing states', \emph{i.e.} a system which reaches these states can never leave them. Well-behavedness at the boundaries for the more general equation \eqref{nD_eqn_for_frequencies} can be checked directly, and has been carried out in Appendix \ref{App_well_behaved}.

>>>>

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' (Also see equation \eqref{nD_stoch_RM_noise_term_alt_representation} in Appendix \ref{App_alternative_representation} for a more elegant representation of these noise terms as an integral against a single Wiener process). For the case $m=2, \mu = 0$ (two interacting types of individuals with no mutations between types), after using the representation of noise terms presented in Appendix \ref{App_alternative_representation} and letting $p = p_1$, equation \eqref{nD_eqn_for_frequencies} becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\sqrt{p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]}dW_t
	\end{aligned}
\end{equation}
In this equation, if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. To see this, one can observe that the RHS of equation \eqref{two_species_stoch_RM} identically vanishes at both $p=0$ and $p=1$, since every term on the RHS contains the product $p(1-p)$. Further, since there is no mutation between types, $p=0$ (extinction of type 1 organisms) and $p=1$ (extinction of type 2 organisms) are both `absorbing states', \emph{i.e.} a system which reaches these states can never leave them. Well-behavedness at the boundaries for the more general equation \eqref{nD_eqn_for_frequencies} can be checked directly, and has been carried out in Appendix \ref{App_well_behaved}.

### 2023-05-01 ind_files/99_Appendix_E_lemma.tex
Correct minor typos

Using this lemma, we can now calculate the stochastic integral terms of our equations. For equation \eqref{nD_stochastic_RM}, the stochastic analog of the replicator-mutator equation, we can use this lemma and the functional form $A_i^+ = x_i\tau_i(\mathbf{x}) + \mu Q_i(\mathbf{x})$ to find that the noise term can be written as a stochastic integral against a single Wiener process $W_t$ is given by
\begin{equation}
\label{nD_stoch_RM_noise_term_alt_representation}
\frac{1}{\sqrt{KN_{K}(t)}}\left[p_i(1-p_i)\tau_i + p_i^2\left(\sum\limits_{j\neq i}\tau_j p_j\right) + \mu\left\{(1-p_i)^2Q_i(\mathbf{p}) + p_i^2 \sum\limits_{j\neq i}Q_j(\mathbf{p})\right\}\right]^{1/2}dW_t
\end{equation}
For equation \eqref{nD_stochastic_Price}, the stochastic analog of the Price equation, we have:
\begin{equation}
dW_{\overline{f}} = \sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)\sqrt{A_i^+}dW_{t}^{(i)} = \left(\sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)^2A_i^+\right)^{1/2}dW_t
\end{equation}
where $dW_t$ is now a single one-dimensional Wiener process. This is precisely the term calculated in equation \eqref{intermediate_2_for_variances} (barring the $1/KN_K^2$ pre-factor), and thus the stochastic term for the mean value is given by:
\begin{equation}
	\label{nD_stoch_Price_noise_term_alt_representation}
	dW_{\overline{f}} = \sqrt{N_K(t) \left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)}dW_t
\end{equation}

>>>>

Using this lemma, we can now calculate the stochastic integral terms of our equations. For equation \eqref{nD_stochastic_RM}, the stochastic analog of the replicator-mutator equation, we can use this lemma and the functional form $A_i^+ = x_i\tau_i(\mathbf{x}) + \mu Q_i(\mathbf{x})$ to find that the noise term can be written as a stochastic integral against a single Wiener process $W_t$ as
\begin{equation}
\label{nD_stoch_RM_noise_term_alt_representation}
\frac{1}{\sqrt{KN_{K}(t)}}\left[p_i(1-p_i)\tau_i + p_i^2\left(\sum\limits_{j\neq i}\tau_j p_j\right) + \mu\left\{(1-p_i)^2Q_i(\mathbf{p}) + p_i^2 \sum\limits_{j\neq i}Q_j(\mathbf{p})\right\}\right]^{1/2}dW_t
\end{equation}
For equation \eqref{nD_stochastic_Price}, the stochastic analog of the Price equation, we have:
\begin{equation}
dW_{\overline{f}} = \sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)\sqrt{A_i^+}dW_{t}^{(i)} = \left(\sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)^2A_i^+\right)^{1/2}dW_t
\end{equation}
where $dW_t$ is now a single one-dimensional Wiener process. This is precisely the term calculated in equation \eqref{intermediate_2_for_variances} (barring the $1/KN_K^2$ pre-factor), and thus the stochastic term for the mean value is given by:
\begin{equation}
	\label{nD_stoch_Price_noise_term_alt_representation}
	dW_{\overline{f}} = \sqrt{N_K(t) \left(\textrm{Cov}(\tau,\left(f - \overline{f}\right)^2) + \overline{\tau}\sigma^2_{f} +\mu \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2Q_i(\mathbf{p})\right)}dW_t
\end{equation}
